### 2019-03-21 index.Rmd
setup for contrast

---
title: ""
output: html_document
---

<br>

![](./meta.png)  
<br>

###A contrast of two meta R packages   
There is extensive support and choice in R to support meta-analyses.  Two common packages in the natural sciences include meta and metafor.  Here, a brief contrast of the strengths of each is described for the synthesis scientist.  Meta is a direct, intuitive choice for rapid implementation of general meta-analytical statistics.  Metafor is a comprehensive package for analyses if the fit models are more complex.  Both packages provide estimates of heterogeneity, excellent visualization tools, and functions to explore publication bias.  Preference and critical outcomes can facilitate choice between these two specific options.  Nonetheless, metafor has a steeper learning curve but greater rewards.

Herein, a quick scrape and analysis of the two R package manuals was done to examine relative complexity of each package. This assumes that every object and function loaded within a package is described in the CRAN manual.

###Data  
```{r, data, warning=FALSE, message=FALSE}
library(tidyverse)
data <- read_csv(url("https://ndownloader.figshare.com/files/14666636"))
#knitr::kable(data)
data

```

### 2019-03-21 lortie_meta/lortie_meta.Rmd
peerj preprint

---
title: A contrast of meta and metafor packages for meta-analyses in R
preprint: true
author: 
  - name: Christopher J. Lortie
    affiliation: 1
    corresponding: true
    email: chris@christopherlortie.info

affiliation:
  - code: 1
    address: NCEAS, UCSB, Santa Barbara, CA. 93101

abstract: >
  There is extensive support and choice in R to support meta-analyses.  Two common packages in the natural sciences include meta and metafor.  Here, a brief contrast of the strengths of each is described for the synthesis scientist.  Meta is a direct, intuitive choice for rapid implementation of general meta-analytical statistics.  Metafor is a comprehensive package for analyses if the fit models are more complex.  Both packages provide estimates of heterogeneity, excellent visualization tools, and functions to explore publication bias.  Preference and critical outcomes can facilitate choice between these two specific options.  Nonetheless, metafor has a steeper learning curve but greater rewards.
header-includes: >
  \usepackage{lipsum}
bibliography: references.bib
output: 
  rticles::peerj_article:
    base_format: rmarkdown::pdf_document # bookdown::pdf_document2 # for using \@ref()
---

# Introduction {-}

Meta-analyses are common and powerful synthesis tools in science.  Typically in the natural sciences, meta-analyses are used as a mechanism to describe and aggregate quantitative evidence from a set of peer-reviewed, primary research publications [@RN3629, @RN4850].  A meta-analysis in the natural sciences, i.e. outside the health sciences, is comprised of a formalized systematic review and analysis of effect sizes and is termed a meta-analysis when statistics examining intervention efficacy are included [@RN3216].  In other fields, the terms systematic review and meta-analysis are used more interchangeable, and meta-statistics are often done on compiled randomized controlled trials or other relatively large datasets in addition to sets of data derived from peer-reviewed publications.  The derived data in the natural sciences (as the intended primary audience here) routinely extracts only the mean values or single point measures from publications [@RN4861].  The synthesis statistics were commonly done using MetaWin [@RN6194] or other GUI-based applications for a number of years.  More recently however, statistics in the fields of ecology and evolution for instance have increasingly moved to the programming language R [@RN6098], and synthesis statistics are no exception.  At least two R packages have risen to prominence for general meta-analytical statistics in the natural sciences - namely meta [@RN6176] and metafor [@RN6175].  Given that meta-analyses are also increasingly published in these same fields [@RN2189, @RN3629], a brief comment on the ecosystem of analytical choices that R provides is beneficial and timely.  We need synthesis to inform evidence-based decisioning, and meta-analyses can be the primary tool if aggregated primary datasets are unavailable.  Furthermore even with primary data in hand, data reduction to effect sizes within primary and synthesis studies is a mechanism to illustrate differences and strength of effects.  These approaches provide the capacity for higher-order analyses and reuse [@RN4835] suggesting that familiarity with effect sizes is both germane and practical.

# The R ecosystem for meta-analyses {-}
Like many fundamental challenges in science, the R developer community provides potential solution sets distributed across multiple packages for synthesis.  Broadly speaking, alternative packages in R sometimes examine an issue from different perspectives and provide unique functions.  In other instances, packages can be very similar or analogs in terms of functionality and use conceptually aligned functions that differ only in nomenclature or arguments.  Scientific synthesists that choose to do a meta-analyses in R have options.  A total of 63 packages associated with various aspects of conducting a meta-analysis have been identified in a comprehensive review and typology of options [@RN6178].  Both meta [@RN6176] and metafor [@RN6175] are amongst 11 generic packages identified [@RN6178].  These two packages are analogs but with different inherent workflows.  There is also rmeta for simple fixed and random effects meta-analyses [@RN6195], mada for diagnostics [@RN6196], [@RN6196], netmeta for frequentist network meta statistics [@RN6197], and mvmeta for multivariate derived data aggregations [@RN6198] to name a few options.  The latter three packages listed have distinct and specific niches for analysis whilst meta and metafor overlap considerably.  Consequently, a brief contrast in facilitating choice between these two packages for the general analyst is provided here.

# Contrast of meta versus metafor {-}
Meta is a well-maintained, recently updated CRAN R package (Version 4.9-4 updated on January 3, 2019) with 31 unique functions, 7 sample datasets, and a reference manual.  There is also an exceptional textbook devoted to meta-analysis in R that focuses primarily on this package [@RN6199].  It is highly capable of resolving most general meta-analytical challenges that an analyst will face including the capacity to include Empirical Bayes estimators as arguments in some functions, predictive meta-statistics, interaction terms, meta-regression, and modifiers.  Note that for some of these methods, the rma.uni function is sourced internally from the package metafor.  This is intriguing but mostly opaque and inconsequential to the user if she prefers the structure of the arguments within functions, the semantics, or the workflow of the meta package.  The primary strengths include its direct and straightforward implementation with minimal (source) lines of code to do an analysis.  Provided one has secured the derived data from the studies and organized into a dataframe with vectors as each key argument within the main meta-model fitting functions, statistics are simple.  The type of response variable such as mean, continuous, or rate is matched to a specific function call such as metamean, metacont, or metarate.  This is semantically intuitive and encourages good thinking before statistics because it engenders consideration of the data.  The effect size calculation is included in this main function and defaults return the most prevalent effect size measure typically associated with those data, but it can also be specified as an argument.  The primary workflow is thus a single step if the user elects to rely on the internal calculations provided in this package.  Exploration of the model is well articulated with funnel, radial, and forest plots.  Z-scores, significance tests, and heterogeneity statistics are printed in the model summary. Publication bias is also provided as a more in-depth function entitled metabias within this package.  There are two standout functions in this package.  The first is a function entitled metagen, and it is a backup, multipurpose tool so to speak that fits a generic inverse variance meta-analysis.  This a handy tool for user-calculated effect size measures or for exploration of statistical trends with reduced data assumptions.  In some fields, there are specific effect size estimates that this function provides a robust, easy-to-fit capacity for statistics.  The second standout function is bubble.metareg for a quick, visual exploration of the outcome of a meta-regression.  It is useful in contemporary data science to use visualization as a means to understand data [@RN4510], but statistical packages do not always provide the means to easily iterate between statistics or model fitting and visualization.  In summary, excepting unique data or statistical issues, this package is directly implemented and effective.

Metafor is a more comprehensive package in many respects.  This package includes 74 functions, 35 datasets, a vignette [@RN4896], flowchart as secondary vignette (https://cran.r-project.org/web/packages/metafor/vignettes/diagram.pdf), and website (http://www.metafor-project.org/doku.php).  The package was last updated on CRAN in June of 2017 (Version 2.0-0).  The text ‘Meta-analysis with R’ also describes implementation of this package [@RN6199] but to a lesser extent than meta.  The depth of the package metafor provides greater capacities relative to the meta package but does come at the expense of a steeper initial learning curve.  Completing a meta-analysis using this package requires an additional step, i.e. effect sizes must be calculated a priori, not within the model fitting process.  This is facilitated with the standalone function escalc, and it can return a wide range of effect sizes measures. Thus, the two-step process begins with firstly compiling and aggregating the derived dataframe to an effect size table then secondly fitting a model.  The data structure is also a bit more fixed for the model fitting, and the nomenclature for this subset of functions is written to parallel more traditional general linear model fitting from conventional statistics.  This is both a strength and limitation because one must plan the model to fit in advance and learn the function and arguments but it is also an advantage as well because model specification uses the familiar notation of tilde.  Model fitting is based on the type of model in the call such a random or fixed effects and not on the type of the response data as in meta package.  Here, it is more akin to conventional general linear model fitting for those familiar with these functions in R.  If the model is more complex with moderators, then this can be directly included in the model fit here via a mods argument whereas in the meta package the model is updated with moderators in a subsequent step.  This suggests that if moderators or covariates in the main model are likely relevant to the analyses, then metafor is a strong starting point.  The model summary also prints z-scores, significance tests, and two sets of heterogeneity estimates.  Forest and radial plots are also provided as additional functions.  Publication bias statistical estimator functions include trimfill and ranktest.  Standout elements of this package are primarily associated with enhanced model fitting capacities such as the function fitstats that provides log-likelihood estimates and AIC or BIC scores on meta-analysis objects.  This package requires a focus on model fitting, and while there is additional effort in specifying the data at the onset of the workflow, the rewards in subsequent tools to handle models are significant.

# Conclusions {-}

Statistics are sometimes about preferences and thinking styles [@RN6087], and scientific synthesis is both an art and a science [@RN3629].  Trade-offs are also common in adopting one ecosystem, analysis tool, or specific package for data wrangling and analyses.  If more rapid, less specified, general meta-analyses are the goal – the package meta is a direct means to an end.  Moderators are added post hoc in additional, update model steps, but the first model fit is a single, intuitive process.  Meta-regression is viable and interaction terms can be included.  The generic meta-analysis function is a superb tool.  Metafor requires the effect size compilation a priori and is thus a bit more coding to prepare for the meta-model.  However, deeper and more complex model fits are inherent in the semantics of these functions.  If the synthesist does have not effect size measure in hand or wishes to calculate effect sizes measures but not for meta-models, the escalc function is invaluable in this package.  In summary, both packages provide the capacity for basic and advanced meta-analyses but more advanced modelling is likely worth the commitment to metafor.

# Acknowledgments {-}  
CJL is a Senior Research Associate at NCEAS and in-kind computational support is provided.

# References

### 2019-03-21 lortie_meta/lortie_meta.tex
peerj preprint

--- 2019-03-21 lortie_meta/lortie_meta.Rmd
@@
----
-title: A contrast of meta and metafor packages for meta-analyses in R
-preprint: true
-author: 
-  - name: Christopher J. Lortie
-    affiliation: 1
-    corresponding: true
-    email: chris@christopherlortie.info
+%% Submissions for peer-review must enable line-numbering
+%% using the lineno option in the \documentclass command.
+%%
+%% Preprints and camera-ready submissions do not need
+%% line numbers, and should have this option removed.
+%%
+%% Please note that the line numbering option requires
+%% version 1.1 or newer of the wlpeerj.cls file, and
+%% the corresponding author info requires v1.2
 
-affiliation:
-  - code: 1
-    address: NCEAS, UCSB, Santa Barbara, CA. 93101
+\documentclass[fleqn,10pt]{wlpeerj} % for preprint submissions
 
-abstract: >
-  There is extensive support and choice in R to support meta-analyses.
-Two common packages in the natural sciences include meta and metafor.
-Here, a brief contrast of the strengths of each is described for the synthesis scientist.
-Meta is a direct, intuitive choice for rapid implementation of general meta-analytical statistics.
-Metafor is a comprehensive package for analyses if the fit models are more complex.
-Both packages provide estimates of heterogeneity, excellent visualization tools, and functions to explore publication bias.
-Preference and critical outcomes can facilitate choice between these two specific options.
-Nonetheless, metafor has a steeper learning curve but greater rewards.
-header-includes: >
+% ZNK -- Adding headers for pandoc
+
+\setlength{\emergencystretch}{3em}
+\providecommand{\tightlist}{
+\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 \usepackage{lipsum}
-bibliography: references.bib
-output: 
-  rticles::peerj_article:
-    base_format: rmarkdown::pdf_document # bookdown::pdf_document2 # for using \@ref()
----
+\usepackage[unicode=true]{hyperref}
+\usepackage{longtable}
 
-# Introduction {-}
+\usepackage{lipsum}
+
+\title{A contrast of meta and metafor packages for meta-analyses in R}
+
+\author[1]{Christopher J. Lortie}
+
+\corrauthor[1]{Christopher J. Lortie}{\href{mailto:chris@christopherlortie.info}{\nolinkurl{chris@christopherlortie.info}}}
+
+\affil[1]{NCEAS, UCSB, Santa Barbara, CA. 93101}
+
+%
+% \author[1]{First Author}
+% \author[2]{Second Author}
+% \affil[1]{Address of first author}
+% \affil[2]{Address of second author}
+% \corrauthor[1]{First Author}{f.author@email.com}
+
+%
+
+\begin{abstract}
+There is extensive support and choice in R to support meta-analyses. Two
+common packages in the natural sciences include meta and metafor. Here,
+a brief contrast of the strengths of each is described for the synthesis
+scientist. Meta is a direct, intuitive choice for rapid implementation
+of general meta-analytical statistics. Metafor is a comprehensive
+package for analyses if the fit models are more complex. Both packages
+provide estimates of heterogeneity, excellent visualization tools, and
+functions to explore publication bias. Preference and critical outcomes
+can facilitate choice between these two specific options. Nonetheless,
+metafor has a steeper learning curve but greater rewards.
+% Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+Dummy abstract text.
+\end{abstract}
+
+\begin{document}
+
+\flushbottom
+\maketitle
+\thispagestyle{empty}
+
+\section*{Introduction}\label{introduction}
+\addcontentsline{toc}{section}{Introduction}
 
 Meta-analyses are common and powerful synthesis tools in science.
-Typically in the natural sciences, meta-analyses are used as a mechanism to describe and aggregate quantitative evidence from a set of peer-reviewed, primary research publications [@RN3629, @RN4850].
-A meta-analysis in the natural sciences, i.e.
-outside the health sciences, is comprised of a formalized systematic review and analysis of effect sizes and is termed a meta-analysis when statistics examining intervention efficacy are included [@RN3216].
-In other fields, the terms systematic review and meta-analysis are used more interchangeable, and meta-statistics are often done on compiled randomized controlled trials or other relatively large datasets in addition to sets of data derived from peer-reviewed publications.
-The derived data in the natural sciences (as the intended primary audience here) routinely extracts only the mean values or single point measures from publications [@RN4861].
-The synthesis statistics were commonly done using MetaWin [@RN6194] or other GUI-based applications for a number of years.
-More recently however, statistics in the fields of ecology and evolution for instance have increasingly moved to the programming language R [@RN6098], and synthesis statistics are no exception.
-At least two R packages have risen to prominence for general meta-analytical statistics in the natural sciences - namely meta [@RN6176] and metafor [@RN6175].
-Given that meta-analyses are also increasingly published in these same fields [@RN2189, @RN3629], a brief comment on the ecosystem of analytical choices that R provides is beneficial and timely.
-We need synthesis to inform evidence-based decisioning, and meta-analyses can be the primary tool if aggregated primary datasets are unavailable.
-Furthermore even with primary data in hand, data reduction to effect sizes within primary and synthesis studies is a mechanism to illustrate differences and strength of effects.
-These approaches provide the capacity for higher-order analyses and reuse [@RN4835] suggesting that familiarity with effect sizes is both germane and practical.
+Typically in the natural sciences, meta-analyses are used as a mechanism
+to describe and aggregate quantitative evidence from a set of
+peer-reviewed, primary research publications (Christopher J. Lortie and
+Bonte 2016, Nakagawa et al. (2017)). A meta-analysis in the natural
+sciences, i.e.~outside the health sciences, is comprised of a formalized
+systematic review and analysis of effect sizes and is termed a
+meta-analysis when statistics examining intervention efficacy are
+included (C. J. Lortie 2014). In other fields, the terms systematic
+review and meta-analysis are used more interchangeable, and
+meta-statistics are often done on compiled randomized controlled trials
+or other relatively large datasets in addition to sets of data derived
+from peer-reviewed publications. The derived data in the natural
+sciences (as the intended primary audience here) routinely extracts only
+the mean values or single point measures from publications (Stewart and
+Schmid 2015). The synthesis statistics were commonly done using MetaWin
+(Rosenberg, Adams, and Gurevitch 2000) or other GUI-based applications
+for a number of years. More recently however, statistics in the fields
+of ecology and evolution for instance have increasingly moved to the
+programming language R (Lai et al. 2019), and synthesis statistics are
+no exception. At least two R packages have risen to prominence for
+general meta-analytical statistics in the natural sciences - namely meta
+(Schwarzer 2019) and metafor (Viechtbauer 2017). Given that
+meta-analyses are also increasingly published in these same fields
+(Cadotte, Mehrkens, and Menge 2012, Christopher J. Lortie and Bonte
+(2016)), a brief comment on the ecosystem of analytical choices that R
+provides is beneficial and timely. We need synthesis to inform
+evidence-based decisioning, and meta-analyses can be the primary tool if
+aggregated primary datasets are unavailable. Furthermore even with
+primary data in hand, data reduction to effect sizes within primary and
+synthesis studies is a mechanism to illustrate differences and strength
+of effects. These approaches provide the capacity for higher-order
+analyses and reuse (Gerstner et al. 2017) suggesting that familiarity
+with effect sizes is both germane and practical.
 
-# The R ecosystem for meta-analyses {-}
-Like many fundamental challenges in science, the R developer community provides potential solution sets distributed across multiple packages for synthesis.
-Broadly speaking, alternative packages in R sometimes examine an issue from different perspectives and provide unique functions.
-In other instances, packages can be very similar or analogs in terms of functionality and use conceptually aligned functions that differ only in nomenclature or arguments.
-Scientific synthesists that choose to do a meta-analyses in R have options.
-A total of 63 packages associated with various aspects of conducting a meta-analysis have been identified in a comprehensive review and typology of options [@RN6178].
-Both meta [@RN6176] and metafor [@RN6175] are amongst 11 generic packages identified [@RN6178].
-These two packages are analogs but with different inherent workflows.
-There is also rmeta for simple fixed and random effects meta-analyses [@RN6195], mada for diagnostics [@RN6196], [@RN6196], netmeta for frequentist network meta statistics [@RN6197], and mvmeta for multivariate derived data aggregations [@RN6198] to name a few options.
-The latter three packages listed have distinct and specific niches for analysis whilst meta and metafor overlap considerably.
-Consequently, a brief contrast in facilitating choice between these two packages for the general analyst is provided here.
+\section*{The R ecosystem for
+meta-analyses}\label{the-r-ecosystem-for-meta-analyses}
+\addcontentsline{toc}{section}{The R ecosystem for meta-analyses}
 
-# Contrast of meta versus metafor {-}
-Meta is a well-maintained, recently updated CRAN R package (Version 4.9-4 updated on January 3, 2019) with 31 unique functions, 7 sample datasets, and a reference manual.
-There is also an exceptional textbook devoted to meta-analysis in R that focuses primarily on this package [@RN6199].
-It is highly capable of resolving most general meta-analytical challenges that an analyst will face including the capacity to include Empirical Bayes estimators as arguments in some functions, predictive meta-statistics, interaction terms, meta-regression, and modifiers.
-Note that for some of these methods, the rma.uni function is sourced internally from the package metafor.
-This is intriguing but mostly opaque and inconsequential to the user if she prefers the structure of the arguments within functions, the semantics, or the workflow of the meta package.
-The primary strengths include its direct and straightforward implementation with minimal (source) lines of code to do an analysis.
-Provided one has secured the derived data from the studies and organized into a dataframe with vectors as each key argument within the main meta-model fitting functions, statistics are simple.
-The type of response variable such as mean, continuous, or rate is matched to a specific function call such as metamean, metacont, or metarate.
-This is semantically intuitive and encourages good thinking before statistics because it engenders consideration of the data.
-The effect size calculation is included in this main function and defaults return the most prevalent effect size measure typically associated with those data, but it can also be specified as an argument.
-The primary workflow is thus a single step if the user elects to rely on the internal calculations provided in this package.
-Exploration of the model is well articulated with funnel, radial, and forest plots.
-Z-scores, significance tests, and heterogeneity statistics are printed in the model summary.
-Publication bias is also provided as a more in-depth function entitled metabias within this package.
-There are two standout functions in this package.
-The first is a function entitled metagen, and it is a backup, multipurpose tool so to speak that fits a generic inverse variance meta-analysis.
-This a handy tool for user-calculated effect size measures or for exploration of statistical trends with reduced data assumptions.
-In some fields, there are specific effect size estimates that this function provides a robust, easy-to-fit capacity for statistics.
-The second standout function is bubble.metareg for a quick, visual exploration of the outcome of a meta-regression.
-It is useful in contemporary data science to use visualization as a means to understand data [@RN4510], but statistical packages do not always provide the means to easily iterate between statistics or model fitting and visualization.
-In summary, excepting unique data or statistical issues, this package is directly implemented and effective.
+Like many fundamental challenges in science, the R developer community
+provides potential solution sets distributed across multiple packages
+for synthesis. Broadly speaking, alternative packages in R sometimes
+examine an issue from different perspectives and provide unique
+functions. In other instances, packages can be very similar or analogs
+in terms of functionality and use conceptually aligned functions that
+differ only in nomenclature or arguments. Scientific synthesists that
+choose to do a meta-analyses in R have options. A total of 63 packages
+associated with various aspects of conducting a meta-analysis have been
+identified in a comprehensive review and typology of options (Polanin,
+Hennessy, and Tanner-Smith 2016). Both meta (Schwarzer 2019) and metafor
+(Viechtbauer 2017) are amongst 11 generic packages identified (Polanin,
+Hennessy, and Tanner-Smith 2016). These two packages are analogs but
+with different inherent workflows. There is also rmeta for simple fixed
+and random effects meta-analyses (Lumley 2018), mada for diagnostics
+(Doebler 2017), (Doebler 2017), netmeta for frequentist network meta
+statistics (Rucker et al. 2019), and mvmeta for multivariate derived
+data aggregations (Gasparrini 2018) to name a few options. The latter
+three packages listed have distinct and specific niches for analysis
+whilst meta and metafor overlap considerably. Consequently, a brief
+contrast in facilitating choice between these two packages for the
+general analyst is provided here.
 
-Metafor is a more comprehensive package in many respects.
-This package includes 74 functions, 35 datasets, a vignette [@RN4896], flowchart as secondary vignette (https://cran.r-project.org/web/packages/metafor/vignettes/diagram.pdf), and website (http://www.metafor-project.org/doku.php).
-The package was last updated on CRAN in June of 2017 (Version 2.0-0).
-The text ‘Meta-analysis with R’ also describes implementation of this package [@RN6199] but to a lesser extent than meta.
-The depth of the package metafor provides greater capacities relative to the meta package but does come at the expense of a steeper initial learning curve.
-Completing a meta-analysis using this package requires an additional step, i.e.
-effect sizes must be calculated a priori, not within the model fitting process.
-This is facilitated with the standalone function escalc, and it can return a wide range of effect sizes measures.
-Thus, the two-step process begins with firstly compiling and aggregating the derived dataframe to an effect size table then secondly fitting a model.
-The data structure is also a bit more fixed for the model fitting, and the nomenclature for this subset of functions is written to parallel more traditional general linear model fitting from conventional statistics.
-This is both a strength and limitation because one must plan the model to fit in advance and learn the function and arguments but it is also an advantage as well because model specification uses the familiar notation of tilde.
-Model fitting is based on the type of model in the call such a random or fixed effects and not on the type of the response data as in meta package.
-Here, it is more akin to conventional general linear model fitting for those familiar with these functions in R.
-If the model is more complex with moderators, then this can be directly included in the model fit here via a mods argument whereas in the meta package the model is updated with moderators in a subsequent step.
-This suggests that if moderators or covariates in the main model are likely relevant to the analyses, then metafor is a strong starting point.
-The model summary also prints z-scores, significance tests, and two sets of heterogeneity estimates.
+\section*{Contrast of meta versus
+metafor}\label{contrast-of-meta-versus-metafor}
+\addcontentsline{toc}{section}{Contrast of meta versus metafor}
+
+Meta is a well-maintained, recently updated CRAN R package (Version
+4.9-4 updated on January 3, 2019) with 31 unique functions, 7 sample
+datasets, and a reference manual. There is also an exceptional textbook
+devoted to meta-analysis in R that focuses primarily on this package
+(Schwarzer, Carpenter, and Rücker 2015). It is highly capable of
+resolving most general meta-analytical challenges that an analyst will
+face including the capacity to include Empirical Bayes estimators as
+arguments in some functions, predictive meta-statistics, interaction
+terms, meta-regression, and modifiers. Note that for some of these
+methods, the rma.uni function is sourced internally from the package
+metafor. This is intriguing but mostly opaque and inconsequential to the
+user if she prefers the structure of the arguments within functions, the
+semantics, or the workflow of the meta package. The primary strengths
+include its direct and straightforward implementation with minimal
+(source) lines of code to do an analysis. Provided one has secured the
+derived data from the studies and organized into a dataframe with
+vectors as each key argument within the main meta-model fitting
+functions, statistics are simple. The type of response variable such as
+mean, continuous, or rate is matched to a specific function call such as
+metamean, metacont, or metarate. This is semantically intuitive and
+encourages good thinking before statistics because it engenders
+consideration of the data. The effect size calculation is included in
+this main function and defaults return the most prevalent effect size
+measure typically associated with those data, but it can also be
+specified as an argument. The primary workflow is thus a single step if
+the user elects to rely on the internal calculations provided in this
+package. Exploration of the model is well articulated with funnel,
+radial, and forest plots. Z-scores, significance tests, and
+heterogeneity statistics are printed in the model summary. Publication
+bias is also provided as a more in-depth function entitled metabias
+within this package. There are two standout functions in this package.
+The first is a function entitled metagen, and it is a backup,
+multipurpose tool so to speak that fits a generic inverse variance
+meta-analysis. This a handy tool for user-calculated effect size
+measures or for exploration of statistical trends with reduced data
+assumptions. In some fields, there are specific effect size estimates
+that this function provides a robust, easy-to-fit capacity for
+statistics. The second standout function is bubble.metareg for a quick,
+visual exploration of the outcome of a meta-regression. It is useful in
+contemporary data science to use visualization as a means to understand
+data (C. Lortie 2017), but statistical packages do not always provide
+the means to easily iterate between statistics or model fitting and
+visualization. In summary, excepting unique data or statistical issues,
+this package is directly implemented and effective.
+
+Metafor is a more comprehensive package in many respects. This package
+includes 74 functions, 35 datasets, a vignette (Viechtbauer 2010),
+flowchart as secondary vignette
+(\url{https://cran.r-project.org/web/packages/metafor/vignettes/diagram.pdf}),
+and website (\url{http://www.metafor-project.org/doku.php}). The package
+was last updated on CRAN in June of 2017 (Version 2.0-0). The text
+`Meta-analysis with R' also describes implementation of this package
+(Schwarzer, Carpenter, and Rücker 2015) but to a lesser extent than
+meta. The depth of the package metafor provides greater capacities
+relative to the meta package but does come at the expense of a steeper
+initial learning curve. Completing a meta-analysis using this package
+requires an additional step, i.e.~effect sizes must be calculated a
+priori, not within the model fitting process. This is facilitated with
+the standalone function escalc, and it can return a wide range of effect
+sizes measures. Thus, the two-step process begins with firstly compiling
+and aggregating the derived dataframe to an effect size table then
+secondly fitting a model. The data structure is also a bit more fixed
+for the model fitting, and the nomenclature for this subset of functions
+is written to parallel more traditional general linear model fitting
+from conventional statistics. This is both a strength and limitation
+because one must plan the model to fit in advance and learn the function
+and arguments but it is also an advantage as well because model
+specification uses the familiar notation of tilde. Model fitting is
+based on the type of model in the call such a random or fixed effects
+and not on the type of the response data as in meta package. Here, it is
+more akin to conventional general linear model fitting for those
+familiar with these functions in R. If the model is more complex with
+moderators, then this can be directly included in the model fit here via
+a mods argument whereas in the meta package the model is updated with
+moderators in a subsequent step. This suggests that if moderators or
+covariates in the main model are likely relevant to the analyses, then
+metafor is a strong starting point. The model summary also prints
+z-scores, significance tests, and two sets of heterogeneity estimates.
 Forest and radial plots are also provided as additional functions.
-Publication bias statistical estimator functions include trimfill and ranktest.
-Standout elements of this package are primarily associated with enhanced model fitting capacities such as the function fitstats that provides log-likelihood estimates and AIC or BIC scores on meta-analysis objects.
-This package requires a focus on model fitting, and while there is additional effort in specifying the data at the onset of the workflow, the rewards in subsequent tools to handle models are significant.
+Publication bias statistical estimator functions include trimfill and
+ranktest. Standout elements of this package are primarily associated
+with enhanced model fitting capacities such as the function fitstats
+that provides log-likelihood estimates and AIC or BIC scores on
+meta-analysis objects. This package requires a focus on model fitting,
+and while there is additional effort in specifying the data at the onset
+of the workflow, the rewards in subsequent tools to handle models are
+significant.
 
-# Conclusions {-}
+\section*{Conclusions}\label{conclusions}
+\addcontentsline{toc}{section}{Conclusions}
 
-Statistics are sometimes about preferences and thinking styles [@RN6087], and scientific synthesis is both an art and a science [@RN3629].
-Trade-offs are also common in adopting one ecosystem, analysis tool, or specific package for data wrangling and analyses.
-If more rapid, less specified, general meta-analyses are the goal – the package meta is a direct means to an end.
-Moderators are added post hoc in additional, update model steps, but the first model fit is a single, intuitive process.
-Meta-regression is viable and interaction terms can be included.
-The generic meta-analysis function is a superb tool.
-Metafor requires the effect size compilation a priori and is thus a bit more coding to prepare for the meta-model.
-However, deeper and more complex model fits are inherent in the semantics of these functions.
-If the synthesist does have not effect size measure in hand or wishes to calculate effect sizes measures but not for meta-models, the escalc function is invaluable in this package.
-In summary, both packages provide the capacity for basic and advanced meta-analyses but more advanced modelling is likely worth the commitment to metafor.
+Statistics are sometimes about preferences and thinking styles (Hector
+2017), and scientific synthesis is both an art and a science
+(Christopher J. Lortie and Bonte 2016). Trade-offs are also common in
+adopting one ecosystem, analysis tool, or specific package for data
+wrangling and analyses. If more rapid, less specified, general
+meta-analyses are the goal -- the package meta is a direct means to an
+end. Moderators are added post hoc in additional, update model steps,
+but the first model fit is a single, intuitive process. Meta-regression
+is viable and interaction terms can be included. The generic
+meta-analysis function is a superb tool. Metafor requires the effect
+size compilation a priori and is thus a bit more coding to prepare for
+the meta-model. However, deeper and more complex model fits are inherent
+in the semantics of these functions. If the synthesist does have not
+effect size measure in hand or wishes to calculate effect sizes measures
+but not for meta-models, the escalc function is invaluable in this
+package. In summary, both packages provide the capacity for basic and
+advanced meta-analyses but more advanced modelling is likely worth the
+commitment to metafor.
 
-# Acknowledgments {-}  
-CJL is a Senior Research Associate at NCEAS and in-kind computational support is provided.
+\section*{Acknowledgments}\label{acknowledgments}
+\addcontentsline{toc}{section}{Acknowledgments}
 
-# References
+CJL is a Senior Research Associate at NCEAS and in-kind computational
+support is provided.
+
+\section*{References}\label{references}
+\addcontentsline{toc}{section}{References}
+
+\hypertarget{refs}{}
+\hypertarget{ref-RN2189}{}
+Cadotte, M.W., L.R. Mehrkens, and D.N.L. Menge. 2012. ``Gauging the
+Impact of Meta-Analysis on Ecology.'' Journal Article.
+\emph{Evolutionary Ecology} 26: 1153--67.
+
+\hypertarget{ref-RN6196}{}
+Doebler, P. 2017. ``Mada.'' Journal Article. \emph{CRAN}, Version 0.5.8.
+
+\hypertarget{ref-RN6198}{}
+Gasparrini, A. 2018. ``Mvmeta.'' Journal Article. \emph{CRAN}, Version
+0.4.11.
+
+\hypertarget{ref-RN4835}{}
+Gerstner, Katharina, David Moreno-Mateos, Jessica Gurevitch, Michael
+Beckmann, Stephan Kambach, Holly P. Jones, and Ralf Seppelt. 2017.
+``Will Your Paper Be Used in a Meta-Analysis? Make the Reach of Your
+Research Broader and Longer Lasting.'' Journal Article. \emph{Methods in
+Ecology and Evolution} 8 (6): 777--84.
+doi:\href{https://doi.org/10.1111/2041-210X.12758}{10.1111/2041-210X.12758}.
+
+\hypertarget{ref-RN6087}{}
+Hector, A. 2017. \emph{The New Statistics with R}. Book. Oxford: Oxford
+University Press.
+
+\hypertarget{ref-RN6098}{}
+Lai, Jiangshan, Christopher J. Lortie, Robert A. Muenchen, Jian Yang,
+and Keping Ma. 2019. ``Evaluating the Popularity of R in Ecology.''
+Journal Article. \emph{Ecosphere} 10 (1): e02567.
+doi:\href{https://doi.org/10.1002/ecs2.2567}{10.1002/ecs2.2567}.
+
+\hypertarget{ref-RN3216}{}
+Lortie, C. J. 2014. ``Formalized Synthesis Opportunities for Ecology:
+Systematic Reviews and Meta-Analyses.'' Journal Article. \emph{Oikos}
+123: 897--902.
+
+\hypertarget{ref-RN4510}{}
+Lortie, Christopher. 2017. ``R for Data Science.'' Journal Article.
+\emph{Journal of Statistical Software} 77: 1--3.
+doi:\href{https://doi.org/10.18637/jss.v077.b01}{10.18637/jss.v077.b01}.
+
+\hypertarget{ref-RN3629}{}
+Lortie, Christopher J., and Dries Bonte. 2016. ``Zen and the Art of
+Ecological Synthesis.'' Journal Article. \emph{Oikos} 125 (3): 285--87.
+doi:\href{https://doi.org/10.1111/oik.03161}{10.1111/oik.03161}.
+
+\hypertarget{ref-RN6195}{}
+Lumley, T. 2018. ``Rmeta.'' Journal Article. \emph{CRAN}, Version 3.0.
+
+\hypertarget{ref-RN4850}{}
+Nakagawa, Shinichi, Daniel W. A. Noble, Alistair M. Senior, and
+Malgorzata Lagisz. 2017. ``Meta-Evaluation of Meta-Analysis: Ten
+Appraisal Questions for Biologists.'' Journal Article. \emph{BMC
+Biology} 15 (1): 18.
+doi:\href{https://doi.org/10.1186/s12915-017-0357-7}{10.1186/s12915-017-0357-7}.
+
+\hypertarget{ref-RN6178}{}
+Polanin, Joshua R., Emily A. Hennessy, and Emily E. Tanner-Smith. 2016.
+``A Review of Meta-Analysis Packages in R.'' Journal Article.
+\emph{Journal of Educational and Behavioral Statistics} 42 (2): 206--42.
+doi:\href{https://doi.org/10.3102/1076998616674315}{10.3102/1076998616674315}.
+
+\hypertarget{ref-RN6194}{}
+Rosenberg, Michael, Dean Adams, and Jessica Gurevitch. 2000.
+\emph{MetaWin: Statistical Software for Meta-Analysis. Version 2.0}.
+Book.
+
+\hypertarget{ref-RN6197}{}
+Rucker, G., U. Krahn, J. Konig, O. Efthimious, and G. Schwarzer. 2019.
+``Netmeta.'' Journal Article. \emph{CRAN} 1.0.1: Version 1.0.1.
+
+\hypertarget{ref-RN6176}{}
+Schwarzer, G. 2019. ``Meta.'' Journal Article. \emph{CRAN}, Version
+4.94.
+
+\hypertarget{ref-RN6199}{}
+Schwarzer, G., J.R. Carpenter, and G. Rücker. 2015. \emph{Meta-Analysis
+with R}. Book. Switzerland: Springer.
+
+\hypertarget{ref-RN4861}{}
+Stewart, G. B., and C. H. Schmid. 2015. ``Lessons from Meta-Analysis in
+Ecology and Evolution: The Need for Trans-Disciplinary Evidence
+Synthesis Methodologies.'' Journal Article. \emph{Research Synthesis
+Methods} 6: 109--10.
+doi:\href{https://doi.org/10.1002/jrsm.1152}{10.1002/jrsm.1152}.
+
+\hypertarget{ref-RN4896}{}
+Viechtbauer, W. 2010. ``Conducting Meta-Analyses in R with the Metafor
+Package.'' Journal Article. \emph{J Stat Software} 36.
+doi:\href{https://doi.org/10.18637/jss.v036.i03}{10.18637/jss.v036.i03}.
+
+\hypertarget{ref-RN6175}{}
+---------. 2017. ``Metafor.'' Journal Article. \emph{CRAN}, Version 2.0.
+
+\end{document}

### 2019-06-05 ms/JEBS/JEBS.Rmd
JEBS prepped

--- 2019-03-21 lortie_meta/lortie_meta.Rmd
@@
 ---
-title: A contrast of meta and metafor packages for meta-analyses in R
-preprint: true
-author: 
-  - name: Christopher J. Lortie
-    affiliation: 1
-    corresponding: true
-    email: chris@christopherlortie.info
+title: "A contrast of meta and metafor packages for meta-analyses in R"
+output: word_document
+bibliography: bibliography.bib
+---
 
-affiliation:
-  - code: 1
-    address: NCEAS, UCSB, Santa Barbara, CA. 93101
-
-abstract: >
+## Abstract  
 There is extensive support and choice in R to support meta-analyses.
@@
 Nonetheless, metafor has a steeper learning curve but greater rewards.
-header-includes: >
-  \usepackage{lipsum}
-bibliography: references.bib
-output: 
-  rticles::peerj_article:
-    base_format: rmarkdown::pdf_document # bookdown::pdf_document2 # for using \@ref()
----
 
-# Introduction {-}
-
+## Introduction
 Meta-analyses are common and powerful synthesis tools in science.
-Typically in the natural sciences, meta-analyses are used as a mechanism to describe and aggregate quantitative evidence from a set of peer-reviewed, primary research publications [@RN3629, @RN4850].
+Typically in the natural sciences, meta-analyses are used as a mechanism to describe and aggregate quantitative evidence from a set of peer-reviewed, primary research publications [@RN3629; @RN4850].
 A meta-analysis in the natural sciences, i.e.
@@
 In other fields, the terms systematic review and meta-analysis are used more interchangeable, and meta-statistics are often done on compiled randomized controlled trials or other relatively large datasets in addition to sets of data derived from peer-reviewed publications.
-The derived data in the natural sciences (as the intended primary audience here) routinely extracts only the mean values or single point measures from publications [@RN4861].
+The derived data in the natural sciences routinely extracts only the mean values or single point measures from publications [@RN4861].
 The synthesis statistics were commonly done using MetaWin [@RN6194] or other GUI-based applications for a number of years.
@@
 At least two R packages have risen to prominence for general meta-analytical statistics in the natural sciences - namely meta [@RN6176] and metafor [@RN6175].
-Given that meta-analyses are also increasingly published in these same fields [@RN2189, @RN3629], a brief comment on the ecosystem of analytical choices that R provides is beneficial and timely.
+Given that meta-analyses are also increasingly published in these same fields [@RN2189; @RN3629], a brief comment on the ecosystem of analytical choices that R provides is beneficial and timely.
 We need synthesis to inform evidence-based decisioning, and meta-analyses can be the primary tool if aggregated primary datasets are unavailable.
@@
 
-# The R ecosystem for meta-analyses {-}
+## The R ecosystem for meta-analyses
 Like many fundamental challenges in science, the R developer community provides potential solution sets distributed across multiple packages for synthesis.
@@
 
-# Contrast of meta versus metafor {-}
+## Contrast of meta versus metafor
 Meta is a well-maintained, recently updated CRAN R package (Version 4.9-4 updated on January 3, 2019) with 31 unique functions, 7 sample datasets, and a reference manual.
@@
 Metafor is a more comprehensive package in many respects.
-This package includes 74 functions, 35 datasets, a vignette [@RN4896], flowchart as secondary vignette (https://cran.r-project.org/web/packages/metafor/vignettes/diagram.pdf), and website (http://www.metafor-project.org/doku.php).
+This package includes 74 functions, 35 datasets, a vignette [@RN4896], [function flowchart as secondary vignette](https://cran.r-project.org/web/packages/metafor/vignettes/diagram.pdf), and [website]( http://www.metafor-project.org/doku.php).
 The package was last updated on CRAN in June of 2017 (Version 2.0-0).
@@
 
-# Conclusions {-}
+## Summary
 
@@
 
-# Acknowledgments {-}  
-CJL is a Senior Research Associate at NCEAS and in-kind computational support is provided.
-
-# References
+## References

### 2019-06-05 ms/R/RJtemplate.tex
JEBS prepped

--- 2019-03-21 lortie_meta/lortie_meta.Rmd
@@
----
-title: A contrast of meta and metafor packages for meta-analyses in R
-preprint: true
-author: 
-  - name: Christopher J. Lortie
-    affiliation: 1
-    corresponding: true
-    email: chris@christopherlortie.info
+% !TeX root = RJwrapper.tex
+\title{A CONTRAST OF META AND METAFOR PACKAGES FOR META-ANALYSES IN R}
+\author{Christopher J. Lortie}
 
-affiliation:
-  - code: 1
-    address: NCEAS, UCSB, Santa Barbara, CA. 93101
+\maketitle
 
-abstract: >
+\abstract{
 There is extensive support and choice in R to support meta-analyses.
@@
 Nonetheless, metafor has a steeper learning curve but greater rewards.
-header-includes: >
-  \usepackage{lipsum}
-bibliography: references.bib
-output: 
-  rticles::peerj_article:
-    base_format: rmarkdown::pdf_document # bookdown::pdf_document2 # for using \@ref()
----
+}
 
-# Introduction {-}
+\section{Introduction}
 
 Meta-analyses are common and powerful synthesis tools in science.
-Typically in the natural sciences, meta-analyses are used as a mechanism to describe and aggregate quantitative evidence from a set of peer-reviewed, primary research publications [@RN3629, @RN4850].
+Typically in the natural sciences, meta-analyses are used as a mechanism to describe and aggregate quantitative evidence from a set of peer-reviewed, primary research publications \citep{RN3629, RN4850}.
 A meta-analysis in the natural sciences, i.e.
-outside the health sciences, is comprised of a formalized systematic review and analysis of effect sizes and is termed a meta-analysis when statistics examining intervention efficacy are included [@RN3216].
+outside the health sciences, is comprised of a formalized systematic review and analysis of effect sizes and is termed a meta-analysis when statistics examining intervention efficacy are included \citep{RN3216}.
 In other fields, the terms systematic review and meta-analysis are used more interchangeable, and meta-statistics are often done on compiled randomized controlled trials or other relatively large datasets in addition to sets of data derived from peer-reviewed publications.
-The derived data in the natural sciences (as the intended primary audience here) routinely extracts only the mean values or single point measures from publications [@RN4861].
-The synthesis statistics were commonly done using MetaWin [@RN6194] or other GUI-based applications for a number of years.
-More recently however, statistics in the fields of ecology and evolution for instance have increasingly moved to the programming language R [@RN6098], and synthesis statistics are no exception.
-At least two R packages have risen to prominence for general meta-analytical statistics in the natural sciences - namely meta [@RN6176] and metafor [@RN6175].
-Given that meta-analyses are also increasingly published in these same fields [@RN2189, @RN3629], a brief comment on the ecosystem of analytical choices that R provides is beneficial and timely.
+The derived data in the natural sciences (as the intended primary audience here) routinely extracts only the mean values or single point measures from publications \citep{RN4861}.
+The synthesis statistics were commonly done using MetaWin \citep{RN6194} or other GUI-based applications for a number of years.
+More recently however, statistics in the fields of ecology and evolution for instance have increasingly moved to the programming language R \citep{RN6098}, and synthesis statistics are no exception.
+At least two R packages have risen to prominence for general meta-analytical statistics in the natural sciences - namely meta \citep{RN6176} and metafor \citep{RN6175}.
+Given that meta-analyses are also increasingly published in these same fields \citep{RN2189, RN3629}, a brief comment on the ecosystem of analytical choices that R provides is beneficial and timely.
 We need synthesis to inform evidence-based decisioning, and meta-analyses can be the primary tool if aggregated primary datasets are unavailable.
 Furthermore even with primary data in hand, data reduction to effect sizes within primary and synthesis studies is a mechanism to illustrate differences and strength of effects.
-These approaches provide the capacity for higher-order analyses and reuse [@RN4835] suggesting that familiarity with effect sizes is both germane and practical.
+These approaches provide the capacity for higher-order analyses and reuse \citep{RN4835} suggesting that familiarity with effect sizes is both germane and practical.
 
-# The R ecosystem for meta-analyses {-}
+\section{The R ecosystem for meta-analyses}
 Like many fundamental challenges in science, the R developer community provides potential solution sets distributed across multiple packages for synthesis.
@@
 Scientific synthesists that choose to do a meta-analyses in R have options.
-A total of 63 packages associated with various aspects of conducting a meta-analysis have been identified in a comprehensive review and typology of options [@RN6178].
-Both meta [@RN6176] and metafor [@RN6175] are amongst 11 generic packages identified [@RN6178].
+A total of 63 packages associated with various aspects of conducting a meta-analysis have been identified in a comprehensive review and typology of options \citep{RN6178}.
+Both meta \citep{RN6176} and metafor \citep{RN6175} are amongst 11 generic packages identified \citep{RN6178}.
 These two packages are analogs but with different inherent workflows.
-There is also rmeta for simple fixed and random effects meta-analyses [@RN6195], mada for diagnostics [@RN6196], [@RN6196], netmeta for frequentist network meta statistics [@RN6197], and mvmeta for multivariate derived data aggregations [@RN6198] to name a few options.
+There is also rmeta for simple fixed and random effects meta-analyses \citep{RN6195}, mada for diagnostics \citep{RN6196}, \citep{RN6196}, netmeta for frequentist network meta statistics \citep{RN6197}, and mvmeta for multivariate derived data aggregations \citep{RN6198} to name a few options.
 The latter three packages listed have distinct and specific niches for analysis whilst meta and metafor overlap considerably.
@@
 
-# Contrast of meta versus metafor {-}
+\section{Contrast of meta versus metafor}
 Meta is a well-maintained, recently updated CRAN R package (Version 4.9-4 updated on January 3, 2019) with 31 unique functions, 7 sample datasets, and a reference manual.
-There is also an exceptional textbook devoted to meta-analysis in R that focuses primarily on this package [@RN6199].
+There is also an exceptional textbook devoted to meta-analysis in R that focuses primarily on this package \citep{RN6199}.
 It is highly capable of resolving most general meta-analytical challenges that an analyst will face including the capacity to include Empirical Bayes estimators as arguments in some functions, predictive meta-statistics, interaction terms, meta-regression, and modifiers.
@@
 The second standout function is bubble.metareg for a quick, visual exploration of the outcome of a meta-regression.
-It is useful in contemporary data science to use visualization as a means to understand data [@RN4510], but statistical packages do not always provide the means to easily iterate between statistics or model fitting and visualization.
+It is useful in contemporary data science to use visualization as a means to understand data \citep{RN4510}, but statistical packages do not always provide the means to easily iterate between statistics or model fitting and visualization.
 In summary, excepting unique data or statistical issues, this package is directly implemented and effective.
@@
 Metafor is a more comprehensive package in many respects.
-This package includes 74 functions, 35 datasets, a vignette [@RN4896], flowchart as secondary vignette (https://cran.r-project.org/web/packages/metafor/vignettes/diagram.pdf), and website (http://www.metafor-project.org/doku.php).
+This package includes 74 functions, 35 datasets, a vignette \citep{RN4896}, \href{https://cran.r-project.org/web/packages/metafor/vignettes/diagram.pdf}{function flowchart as secondary vignette}, and \href{http://www.metafor-project.org/doku.php}{website}.
 The package was last updated on CRAN in June of 2017 (Version 2.0-0).
-The text ‘Meta-analysis with R’ also describes implementation of this package [@RN6199] but to a lesser extent than meta.
+The text ‘Meta-analysis with R’ also describes implementation of this package \citep{RN6199} but to a lesser extent than meta.
 The depth of the package metafor provides greater capacities relative to the meta package but does come at the expense of a steeper initial learning curve.
@@
 
-# Conclusions {-}
+\section{Summary}
 
-Statistics are sometimes about preferences and thinking styles [@RN6087], and scientific synthesis is both an art and a science [@RN3629].
+Statistics are sometimes about preferences and thinking styles \cite{RN6087}, and scientific synthesis is both an art and a science \cite{RN3629}.
 Trade-offs are also common in adopting one ecosystem, analysis tool, or specific package for data wrangling and analyses.
@@
 
-# Acknowledgments {-}  
-CJL is a Senior Research Associate at NCEAS and in-kind computational support is provided.
+\bibliography{RJreferences}
 
-# References
+\address{Christopher J. Lortie\\
+  The National Center for Ecological Analysis and Synthesis, UCSB\\
+  735 State St. Santa Barbara\\
+  USA\\
+  \href{http://orcid.org/0000-0002-4291-7023}{ORCID}\\
+  \email{chris@christopherlortie.info}}

### 2019-06-05 ms/R/cover_letter.Rmd
JEBS prepped

---
title: ""
output:
  pdf_document: default
  html_document: default
---
March 15th, 2019

##Cover letter
###A CONTRAST OF META AND METAFOR PACKAGES FOR META-ANALYSES IN R

Dear Editors,

Meta-analyses are critical scientific synthesis tools.  Their use is increasing exponentially in the natural sciences, and R is a logical (and best choice) to get the statistics done.  In ecology, evolution, and the environmental sciences in particular, two packages dominate the contemporary landscape. A contrast here is provided to guide the synthesist in choosing between these two options. Both excel but in different capacities. This comparison article also highlights the concepts of workflows, semantics, and preference as a set of criteria for R users to consider for any challenge given the incredible volume and range of packages on CRAN.

Warm regards,  
Christopher Lortie

### 2019-06-05 ms/JEBS/cover_letter.Rmd
cover letter added

--- 2019-06-05 ms/R/cover_letter.Rmd
@@
 ---
-March 15th, 2019
+June 5th, 2019
 
-##Cover letter
-###A CONTRAST OF META AND METAFOR PACKAGES FOR META-ANALYSES IN R
+## Cover letter
+### A CONTRAST OF META AND METAFOR PACKAGES FOR META-ANALYSES IN R
 
@@
 Their use is increasing exponentially in the natural sciences, and R is a logical (and best choice) to get the statistics done.
-In ecology, evolution, and the environmental sciences in particular, two packages dominate the contemporary landscape.
+In many of these disciplines, two packages dominate the contemporary landscape.
 A contrast here is provided to guide the synthesist in choosing between these two options.

### 2019-08-06 ms/JSS/softwarereview.tex
setup for pasting code across now

\documentclass[softwarereview]{jss}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% reviewer
\Reviewer{Christopher J. Lortie\\NCEAS}
\Plainreviewer{Christopher Lortie}

%% about the software
\Softwaretitle{\pkg{Meta} 4.9-5}
\Softwaretitle{\pkg{Metafor} 2.1-0}
\Plaintitle{Meta 4.9-5}
\Plaintitle{Meta 4.9-5 and Metafor 2.1-0}

%% if different from \Softwaretitle also set
%% \Shorttitle{Aabel 1.5.7}
\Publisher{CRAN}
\Pubaddress{Vienna, Austria}
\Price{USD~0}
\URL{https://cran.r-project.org/web/packages/metafor/index.html}

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{1}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}

%% address of reviewer
\Address{
  Christopher J. Lortie\\
  University of California Santa Barbara\\
  The National Center for Ecological Analysis and Synthesis\\
  Santa Barbara, CA, USA\\
  E-mail: \email{chris@christopherlortie.info}\\
  URL: \url{http://www.christopherlortie.info}
}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%% include the review as usual
%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.

\end{document}

### 2019-08-07 ms/JSS/softwarereview.tex
review complete

@@
 %% Note that you should use the \pkg{}, \proglang{} and \code{} commands.
+\textbf{Abstract} \newline
+
+There is extensive support and choice in \proglang{R} to support meta-analyses.
+Two common packages in the natural sciences include meta and metafor.
+Here, a brief contrast of the strengths of each is described for the synthesis scientist.
+\pkg{Meta} is a direct, intuitive choice for rapid implementation of general meta-analytical statistics.
+\pkg{Metafor} is a comprehensive package for analyses if the fit models are more complex.
+Both packages provide estimates of heterogeneity, excellent visualization tools, and functions to explore publication bias.
+Preference and critical outcomes can facilitate choice between these two specific options.
+Nonetheless, metafor has a steeper learning curve but greater rewards.
+\newline
+
+\textbf{A contrast of meta and metafor packages for meta-analyses in R} \newline
+
+Meta-analyses are common and powerful synthesis tools in science.
+Typically in the natural sciences, meta-analyses are used as a mechanism to describe and aggregate quantitative evidence from a set of peer-reviewed, primary research publications \citep{RN3629, RN4850}.
+A meta-analysis in the natural sciences, i.e.
+outside the health sciences, is comprised of a formalized systematic review and analysis of effect sizes and is termed a meta-analysis when statistics examining intervention efficacy are included \citep{RN3216}.
+In other fields, the terms systematic review and meta-analysis are used more interchangeable, and meta-statistics are often done on compiled randomized controlled trials or other relatively large datasets in addition to sets of data derived from peer-reviewed publications.
+The derived data in the natural sciences routinely extracts only the mean values or single point measures from publications \citep{RN4861}.
+The synthesis statistics were commonly done using MetaWin \citep{RN6194} or other GUI-based applications for a number of years.
+More recently however, statistics in the fields of ecology and evolution for instance have increasingly moved to the programming language \proglang{R} \citep{RN6098}, and synthesis statistics are no exception.
+At least two R packages have risen to prominence for general meta-analytical statistics in the natural sciences - namely \pkg{Meta} \citep{RN6176} and \pkg{Metafor} \citep{RN6175}.
+Given that meta-analyses are also increasingly published in these same fields \citep{RN2189, RN3629}, a brief comment on the ecosystem of analytical choices that R provides is beneficial and timely.
+We need synthesis to inform evidence-based decisioning, and meta-analyses can be the primary tool if aggregated primary datasets are unavailable.
+Furthermore even with primary data in hand, data reduction to effect sizes within primary and synthesis studies is a mechanism to illustrate differences and strength of effects.
+These approaches provide the capacity for higher-order analyses and reuse \citep{RN4835} suggesting that familiarity with effect sizes is both germane and practical.
+\newline
+
+\textbf{The R ecosystem for meta-analyses} \newline
+Like many fundamental challenges in science, the \proglang{R} developer community provides potential solution sets distributed across multiple packages for synthesis.
+Broadly speaking, alternative packages in \proglang{R} sometimes examine an issue from different perspectives and provide unique functions.
+In other instances, packages can be very similar or analogs in terms of functionality and use conceptually aligned functions that differ only in nomenclature or arguments.
+Scientific synthesists that choose to do a meta-analyses in \proglang{R} have options.
+A total of 63 packages associated with various aspects of conducting a meta-analysis have been identified in a comprehensive review and typology of options \citep{RN6178}.
+Both \pkg{Meta} \citep{RN6176} and \pkg{Metafor} \citep{RN6175} are amongst 11 generic packages identified \citep{RN6178}.
+These two packages are analogs but with different inherent workflows.
+There is also \pkg{rmeta} for simple fixed and random effects meta-analyses \citep{RN6195}, \pkg{mada} for diagnostics \citep{RN6196}, \citep{RN6196}, \pkg{netmeta} for frequentist network meta statistics \citep{RN6197}, and \pkg{mvmeta} for multivariate derived data aggregations \citep{RN6198} to name a few options.
+The latter three packages listed have distinct and specific niches for analysis whilst \pkg{Meta} and \pkg{Metafor} overlap considerably.
+Consequently, a brief contrast in facilitating choice between these two packages for the general analyst is provided here.
+\newline
+
+\textbf{Contrast of meta versus metafor} \newline
+\pkg{Meta} is a well-maintained, recently updated CRAN R package (Version 4.9-4 updated on January 3, 2019) with 31 unique functions, 7 sample datasets, and a reference manual.
+There is also an exceptional textbook devoted to meta-analysis in \proglang{R} that focuses primarily on this package \citep{RN6199}.
+It is highly capable of resolving most general meta-analytical challenges that an analyst will face including the capacity to include Empirical Bayes estimators as arguments in some functions, predictive meta-statistics, interaction terms, meta-regression, and modifiers.
+Note that for some of these methods, the rma.uni function is sourced internally from the package \pkg{Metafor}.
+This is intriguing but mostly opaque and inconsequential to the user if she prefers the structure of the arguments within functions, the semantics, or the workflow of the \pkg{Meta}  package.
+The primary strengths include its direct and straightforward implementation with minimal (source) lines of code to do an analysis.
+Provided one has secured the derived data from the studies and organized into a dataframe with vectors as each key argument within the main meta-model fitting functions, statistics are simple.
+The type of response variable such as mean, continuous, or rate is matched to a specific function call such as metamean, metacont, or metarate.
+This is semantically intuitive and encourages good thinking before statistics because it engenders consideration of the data.
+The effect size calculation is included in this main function and defaults return the most prevalent effect size measure typically associated with those data, but it can also be specified as an argument.
+The primary workflow is thus a single step if the user elects to rely on the internal calculations provided in this package.
+Exploration of the model is well articulated with funnel, radial, and forest plots.
+Z-scores, significance tests, and heterogeneity statistics are printed in the model summary.
+Publication bias is also provided as a more in-depth function entitled metabias within this package.
+There are two standout functions in this package.
+The first is a function entitled metagen, and it is a backup, multipurpose tool so to speak that fits a generic inverse variance meta-analysis.
+This a handy tool for user-calculated effect size measures or for exploration of statistical trends with reduced data assumptions.
+In some fields, there are specific effect size estimates that this function provides a robust, easy-to-fit capacity for statistics.
+The second standout function is bubble.metareg for a quick, visual exploration of the outcome of a meta-regression.
+It is useful in contemporary data science to use visualization as a means to understand data \citep{RN4510}, but statistical packages do not always provide the means to easily iterate between statistics or model fitting and visualization.
+In summary, excepting unique data or statistical issues, this package is directly implemented and effective.
+\newline
+
+\pkg{Metafor} is a more comprehensive package in many respects.
+This package includes 74 functions, 35 datasets, a vignette \citep{RN4896}, function flowchart as secondary vignette, and website.
+The package was last updated on CRAN in June of 2017 (Version 2.0-0).
+The text ‘Meta-analysis with R’ also describes implementation of this package \citep{RN6199} but to a lesser extent than meta.
+The depth of the package \pkg{Metafor} provides greater capacities relative to the \pkg{Meta} package but does come at the expense of a steeper initial learning curve.
+Completing a meta-analysis using this package requires an additional step, i.e.
+effect sizes must be calculated a priori, not within the model fitting process.
+This is facilitated with the standalone function escalc, and it can return a wide range of effect sizes measures.
+Thus, the two-step process begins with firstly compiling and aggregating the derived dataframe to an effect size table then secondly fitting a model.
+The data structure is also a bit more fixed for the model fitting, and the nomenclature for this subset of functions is written to parallel more traditional general linear model fitting from conventional statistics.
+This is both a strength and limitation because one must plan the model to fit in advance and learn the function and arguments but it is also an advantage as well because model specification uses the familiar notation of tilde.
+Model fitting is based on the type of model in the call such a random or fixed effects and not on the type of the response data as in \pkg{Meta} package.
+Here, it is more akin to conventional general linear model fitting for those familiar with these functions in \proglang{R}.
+If the model is more complex with moderators, then this can be directly included in the model fit here via a mods argument whereas in the \pkg{Meta} package the model is updated with moderators in a subsequent step.
+This suggests that if moderators or covariates in the main model are likely relevant to the analyses, then \pkg{Metafor} is a strong starting point.
+The model summary also prints z-scores, significance tests, and two sets of heterogeneity estimates.
+Forest and radial plots are also provided as additional functions.
+Publication bias statistical estimator functions include trimfill and ranktest.
+Standout elements of this package are primarily associated with enhanced model fitting capacities such as the function fitstats that provides log-likelihood estimates and AIC or BIC scores on meta-analysis objects.
+This package requires a focus on model fitting, and while there is additional effort in specifying the data at the onset of the workflow, the rewards in subsequent tools to handle models are significant.
+\newline
+
+\textbf{Summary} \newline
+Statistics are sometimes about preferences and thinking styles \citep{RN6087}, and scientific synthesis is both an art and a science \citep{RN3629}.
+Trade-offs are also common in adopting one ecosystem, analysis tool, or specific package for data wrangling and analyses.
+If more rapid, less specified, general meta-analyses are the goal – the package meta is a direct means to an end.
+Moderators are added post hoc in additional, update model steps, but the first model fit is a single, intuitive process.
+Meta-regression is viable and interaction terms can be included.
+The generic meta-analysis function is a superb tool.
+\pkg{Metafor} requires the effect size compilation a priori and is thus a bit more coding to prepare for the meta-model.
+However, deeper and more complex model fits are inherent in the semantics of these functions.
+If the synthesist does have not effect size measure in hand or wishes to calculate effect sizes measures but not for meta-models, the escalc function is invaluable in this package.
+In summary, both packages provide the capacity for basic and advanced meta-analyses but more advanced modelling is likely worth the commitment to metafor.
+\newline
+\bibliography{lit}
 

### 2020-03-27 ms/2020/contrastRmd.Rmd
tidy up

--- 2019-03-21 lortie_meta/lortie_meta.Rmd
@@
 ---
-title: A contrast of meta and metafor packages for meta-analyses in R
-preprint: true
-author: 
-  - name: Christopher J. Lortie
-    affiliation: 1
-    corresponding: true
-    email: chris@christopherlortie.info
+title: "A contrast of meta and metafor packages for meta-analyses in R"
+output: pdf_document
+bibliography: references.bib
+---
 
-affiliation:
-  - code: 1
-    address: NCEAS, UCSB, Santa Barbara, CA. 93101
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+```
 
-abstract: >
-  There is extensive support and choice in R to support meta-analyses.
+Christopher J. Lortie
+
+Department of Biology, York University  
+Toronto, ON, Canada.  M3J1P3.  
+[lortie@yorku.ca](mailto:lortie@yorku.ca)  
+PH: 416.736.2100 x20588
+
+### Abstract
+There is extensive choice in R to support meta-analyses.
 Two common packages in the natural sciences include meta and metafor.
-Here, a brief contrast of the strengths of each is described for the synthesis scientist.
+A brief contrast of the strengths of each is described here for the synthesis scientist.
 Meta is a direct, intuitive choice for rapid implementation of general meta-analytical statistics.
-Metafor is a comprehensive package for analyses if the fit models are more complex.
+Metafor is a comprehensive package for analyses for relatively more complex fit models.
 Both packages provide estimates of heterogeneity, excellent visualization tools, and functions to explore publication bias.
@@
 Nonetheless, metafor has a steeper learning curve but greater rewards.
-header-includes: >
-  \usepackage{lipsum}
-bibliography: references.bib
-output: 
-  rticles::peerj_article:
-    base_format: rmarkdown::pdf_document # bookdown::pdf_document2 # for using \@ref()
----
 
-# Introduction {-}
+\newpage
 
+### Introduction
 Meta-analyses are common and powerful synthesis tools in science.
@@
 
-# The R ecosystem for meta-analyses {-}
+### The R ecosystem for meta-analyses  
 Like many fundamental challenges in science, the R developer community provides potential solution sets distributed across multiple packages for synthesis.
@@
 
-# Contrast of meta versus metafor {-}
+### Contrast of meta versus metafor  
 Meta is a well-maintained, recently updated CRAN R package (Version 4.9-4 updated on January 3, 2019) with 31 unique functions, 7 sample datasets, and a reference manual.
@@
 
-# Conclusions {-}
-
+### Conclusions  
 Statistics are sometimes about preferences and thinking styles [@RN6087], and scientific synthesis is both an art and a science [@RN3629].
@@
 
-# Acknowledgments {-}  
-CJL is a Senior Research Associate at NCEAS and in-kind computational support is provided.
+\newpage
 
-# References
+### Literature cited

### 2020-03-27 ms/lortie_meta/lortie_meta.Rmd
tidy up

--- 2019-03-21 lortie_meta/lortie_meta.Rmd
@@
 abstract: >
-  There is extensive support and choice in R to support meta-analyses.
+  There is extensive choice in R to support meta-analyses.
 Two common packages in the natural sciences include meta and metafor.
-Here, a brief contrast of the strengths of each is described for the synthesis scientist.
+A brief contrast of the strengths of each is described here for the synthesis scientist.
 Meta is a direct, intuitive choice for rapid implementation of general meta-analytical statistics.
-Metafor is a comprehensive package for analyses if the fit models are more complex.
+Metafor is a comprehensive package for analyses for relatively more complex fit models.
 Both packages provide estimates of heterogeneity, excellent visualization tools, and functions to explore publication bias.

### 2020-03-28 ms/2020/contrastRmd.Rmd
revising abstract

@@
 There is extensive choice in R to support meta-analyses.
-Two common packages in the natural sciences include meta and metafor.
-A brief contrast of the strengths of each is described here for the synthesis scientist.
+Two common packages used in the sciences include meta and metafor.
+A brief contrast of their strengths is described here for the synthesis scientist.
 Meta is a direct, intuitive choice for rapid implementation of general meta-analytical statistics.
-Metafor is a comprehensive package for analyses for relatively more complex fit models.
+Metafor is a comprehensive package best suited for relatively more complex models.
 Both packages provide estimates of heterogeneity, excellent visualization tools, and functions to explore publication bias.

### 2020-03-28 ms/2020/contrastRmd.Rmd
added keywords
edit and then add stata

@@
 
+### Keywords
+contrasts, computational tools, Meta, meta-analyses, Metafor, methods, R, review
+
 \newpage
@@
 Meta-analyses are common and powerful synthesis tools in science.
-Typically in the natural sciences, meta-analyses are used as a mechanism to describe and aggregate quantitative evidence from a set of peer-reviewed, primary research publications [@RN3629, @RN4850].
+Typically in the natural sciences, meta-analyses are used as a mechanism to describe and aggregate quantitative evidence from a set of peer-reviewed, primary research publications [@RN3629; @RN4850].
 A meta-analysis in the natural sciences, i.e.

### 2020-03-29 ms/2020/contrastRmd.Rmd
better!

@@
 Both packages provide estimates of heterogeneity, excellent visualization tools, and functions to explore publication bias.
-Preference and critical outcomes can facilitate choice between these two specific options.
+Preference and specific needs will facilitate choice between these two options.
 Nonetheless, metafor has a steeper learning curve but greater rewards.
@@
 outside the health sciences, is comprised of a formalized systematic review and analysis of effect sizes and is termed a meta-analysis when statistics examining intervention efficacy are included [@RN3216].
-In other fields, the terms systematic review and meta-analysis are used more interchangeable, and meta-statistics are often done on compiled randomized controlled trials or other relatively large datasets in addition to sets of data derived from peer-reviewed publications.
-The derived data in the natural sciences (as the intended primary audience here) routinely extracts only the mean values or single point measures from publications [@RN4861].
-The synthesis statistics were commonly done using MetaWin [@RN6194] or other GUI-based applications for a number of years.
+In other fields, the terms systematic review and meta-analysis are used more interchangeable, and meta-statistics are often done on compiled randomized controlled trials or other relatively large datasets in addition to data derived from peer-reviewed publications.
+The derived data in the natural sciences routinely extracts only the mean values or single point measures from publications [@RN4861].
+The synthesis statistics were commonly done using MetaWin in some fields [@RN6194] or other GUI-based applications for a number of years.
 More recently however, statistics in the fields of ecology and evolution for instance have increasingly moved to the programming language R [@RN6098], and synthesis statistics are no exception.
@@
 Broadly speaking, alternative packages in R sometimes examine an issue from different perspectives and provide unique functions.
-In other instances, packages can be very similar or analogs in terms of functionality and use conceptually aligned functions that differ only in nomenclature or arguments.
+In other instances, packages can be very similar or analogs in terms of functionality and use conceptually aligned functions that differ only in nomenclature or arguments [@RN6507].
 Scientific synthesists that choose to do a meta-analyses in R have options.
@@
 ### Contrast of meta versus metafor  
-Meta is a well-maintained, recently updated CRAN R package (Version 4.9-4 updated on January 3, 2019) with 31 unique functions, 7 sample datasets, and a reference manual.
+Meta is a well-maintained, recently updated CRAN R package (Version 4.11-0 updated on Feb 20, 2020) with 31 unique functions, 7 sample datasets, and a reference manual.
 There is also an exceptional textbook devoted to meta-analysis in R that focuses primarily on this package [@RN6199].
@@
 This package includes 74 functions, 35 datasets, a vignette [@RN4896], flowchart as secondary vignette (https://cran.r-project.org/web/packages/metafor/vignettes/diagram.pdf), and website (http://www.metafor-project.org/doku.php).
-The package was last updated on CRAN in June of 2017 (Version 2.0-0).
+The package was last updated on March 19, 2020 (Version 2.4-0).
 The text ‘Meta-analysis with R’ also describes implementation of this package [@RN6199] but to a lesser extent than meta.
@@
 Thus, the two-step process begins with firstly compiling and aggregating the derived dataframe to an effect size table then secondly fitting a model.
-The data structure is also a bit more fixed for the model fitting, and the nomenclature for this subset of functions is written to parallel more traditional general linear model fitting from conventional statistics.
-This is both a strength and limitation because one must plan the model to fit in advance and learn the function and arguments but it is also an advantage as well because model specification uses the familiar notation of tilde.
+The data structure is also a bit more rigid for the model fitting, and the nomenclature for this subset of functions is written to parallel more traditional general linear model fitting from conventional statistics.
+This is both a strength and limitation because one must plan the model to fit in advance and learn the function and arguments, but it is also an advantage as well because model specification uses the familiar notation of tilde.
 Model fitting is based on the type of model in the call such a random or fixed effects and not on the type of the response data as in meta package.
@@
 This suggests that if moderators or covariates in the main model are likely relevant to the analyses, then metafor is a strong starting point.
-The model summary also prints z-scores, significance tests, and two sets of heterogeneity estimates.
+The model summary also prints Z-scores, significance tests, and two sets of heterogeneity estimates.
 Forest and radial plots are also provided as additional functions.
 Publication bias statistical estimator functions include trimfill and ranktest.
-Standout elements of this package are primarily associated with enhanced model fitting capacities such as the function fitstats that provides log-likelihood estimates and AIC or BIC scores on meta-analysis objects.
+Standout elements of this package include GOSH plots that provide a graphical display of study heterogeneity [@RN6582] and the enhanced model fitting capacities such as the function fitstats that provides log-likelihood estimates and AIC or BIC scores on meta-analysis objects.
 This package requires a focus on model fitting, and while there is additional effort in specifying the data at the onset of the workflow, the rewards in subsequent tools to handle models are significant.

### 2020-03-30 ms/2020/contrastRmd.Rmd
appendix added setup for table 1

@@
 
-Christopher J. Lortie
+Christopher J. Lortie^1,2^ and Alex Filazzola^3^
 
-Department of Biology, York University  
+1. Department of Biology, York University  
 Toronto, ON, Canada.  M3J1P3.  
@@
 PH: 416.736.2100 x20588
+
+2. The National Center for Ecological Analysis and Synthesis,  
+UCSB, California. USA.
+
+3. Biological Sciences, University of Alberta,  
+Edmonton, AB, Canada,
 
@@
 ### Contrast of meta versus metafor  
-Meta is a well-maintained, recently updated CRAN R package (Version 4.11-0 updated on Feb 20, 2020) with 31 unique functions, 7 sample datasets, and a reference manual.
+Meta is a well-maintained, recently updated CRAN R package (Version 4.11-0 updated on Feb 20, 2020) with 31 unique functions, 7 sample datasets (Appendix A), and a reference manual.
 There is also an exceptional textbook devoted to meta-analysis in R that focuses primarily on this package [@RN6199].
@@
 Metafor is a more comprehensive package in many respects.
-This package includes 74 functions, 35 datasets, a vignette [@RN4896], flowchart as secondary vignette (https://cran.r-project.org/web/packages/metafor/vignettes/diagram.pdf), and website (http://www.metafor-project.org/doku.php).
+This package includes 74 functions, 35 datasets (Appendix A), a vignette [@RN4896], flowchart as secondary vignette (https://cran.r-project.org/web/packages/metafor/vignettes/diagram.pdf), and website (http://www.metafor-project.org/doku.php).
 The package was last updated on March 19, 2020 (Version 2.4-0).
@@
 \newpage
+### Table 1.  
+A contrast of meta and metafor using
 
+\newpage
+### Appendix A.  
+A list of objects included in each package when loaded into the R environment.  
+```{r appendix, message=FALSE, warning=FALSE, echo=FALSE}
+library(tidyverse)
+library(kableExtra)
+
+appendix <- read_csv("appendix.csv")
+knitr::kable(appendix, "latex", booktabs = TRUE, longtable = TRUE) %>%
+  column_spec(1, width = "2cm") %>%
+  column_spec(2, width = "4cm") %>%
+  column_spec(3, width = "5cm") %>%
+  column_spec(5, width = "2cm") %>%
+  kable_styling(latex_options = c("hold_position", "repeat_header"))
+```
+
+\newpage
 ### Literature cited

### 2020-03-30 ms/2020/contrastRmd.Rmd
contrast table added

@@
 ### Table 1.  
-A contrast of meta and metafor using
+A contrast of meta and metafor using the critical principles developed in 'A checklist for choosing between R packages in ecology and evolution' [@RN6507].
+
+```{r table1, message=FALSE, warning=FALSE, echo=FALSE}
+library(tidyverse)
+library(kableExtra)
+
+table1 <- read_csv("table1.csv")
+knitr::kable(table1, "latex", booktabs = TRUE, longtable = TRUE) %>%
+  column_spec(1, width = "1cm") %>%
+  column_spec(2, width = "3cm") %>%
+  column_spec(3, width = "5cm") %>%
+  column_spec(4, width = "5cm") %>%
+  kable_styling(latex_options = c("hold_position", "repeat_header"))
+```
 

### 2020-03-30 contrast.Rmd
getting there.

---
title: ""
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### A contrast of meta and metafor
Contrasted meta and metafor using the classic [Cochrane bronchoconstriction dataset]((https://www.cochrane.org/CD001183/AIRWAYS_nedocromil-sodium-for-preventing-exercise-induced-bronchoconstriction)) used in many texts to demonstrate meta-analsysis. [Here is a link to these data as a tidy csv files on figshare.](https://figshare.com/articles/Meta_thinking_data_preventing_exercise-induced_bronchoconstriction/7767020)

#### Meta
```{r meta, warning=FALSE, message=FALSE}
#load primary libraries
library(tidyverse) #tidy functions to read and format data
library(meta) #meta package current version

#Classic bronchoconstriction data from Cochrane
data <- read_csv(url("https://ndownloader.figshare.com/files/14460386"))

#continuous data
m1 <- metacont(Ne, Me, Se, Nc, Mc, Sc, data = data, sm = "SMD")
summary(m1)
forest(m1, layout = "JAMA")
metabias(m1)
funnel(m1)
radial(m1)

#cumulative evidence
forest(metacum(m1, pooled = "random"))

```

### Metafor
```{r, metafor, warning=FALSE, message=FALSE}
#load primary libraries
library(tidyverse) #tidy functions to read and format data
library(metafor) #metafor package current version

#Classic bronchoconstriction data from Cochrane
data <- read_csv(url("https://ndownloader.figshare.com/files/14460386"))

```

### 2020-03-30 ms/2020/appendix_a.Rmd
getting there.

--- 2020-03-30 ms/2020/contrastRmd.Rmd
@@
 ---
-title: "A contrast of meta and metafor packages for meta-analyses in R"
+title: "Appendix A"
 output: pdf_document
@@
 
-Christopher J. Lortie^1,2^ and Alex Filazzola^3^
-
-1. Department of Biology, York University  
-Toronto, ON, Canada.  M3J1P3.  
-[lortie@yorku.ca](mailto:lortie@yorku.ca)  
-PH: 416.736.2100 x20588
-
-2. The National Center for Ecological Analysis and Synthesis,  
-UCSB, California. USA.
-
-3. Biological Sciences, University of Alberta,  
-Edmonton, AB, Canada,
-
-### Abstract
-There is extensive choice in R to support meta-analyses.
-Two common packages used in the sciences include meta and metafor.
-A brief contrast of their strengths is described here for the synthesis scientist.
-Meta is a direct, intuitive choice for rapid implementation of general meta-analytical statistics.
-Metafor is a comprehensive package best suited for relatively more complex models.
-Both packages provide estimates of heterogeneity, excellent visualization tools, and functions to explore publication bias.
-Preference and specific needs will facilitate choice between these two options.
-Nonetheless, metafor has a steeper learning curve but greater rewards.
-
-### Keywords
-contrasts, computational tools, Meta, meta-analyses, Metafor, methods, R, review
-
-\newpage
-
-### Introduction
-Meta-analyses are common and powerful synthesis tools in science.
-Typically in the natural sciences, meta-analyses are used as a mechanism to describe and aggregate quantitative evidence from a set of peer-reviewed, primary research publications [@RN3629; @RN4850].
-A meta-analysis in the natural sciences, i.e.
-outside the health sciences, is comprised of a formalized systematic review and analysis of effect sizes and is termed a meta-analysis when statistics examining intervention efficacy are included [@RN3216].
-In other fields, the terms systematic review and meta-analysis are used more interchangeable, and meta-statistics are often done on compiled randomized controlled trials or other relatively large datasets in addition to data derived from peer-reviewed publications.
-The derived data in the natural sciences routinely extracts only the mean values or single point measures from publications [@RN4861].
-The synthesis statistics were commonly done using MetaWin in some fields [@RN6194] or other GUI-based applications for a number of years.
-More recently however, statistics in the fields of ecology and evolution for instance have increasingly moved to the programming language R [@RN6098], and synthesis statistics are no exception.
-At least two R packages have risen to prominence for general meta-analytical statistics in the natural sciences - namely meta [@RN6176] and metafor [@RN6175].
-Given that meta-analyses are also increasingly published in these same fields [@RN2189; @RN3629], a brief comment on the ecosystem of analytical choices that R provides is beneficial and timely.
-We need synthesis to inform evidence-based decisioning, and meta-analyses can be the primary tool if aggregated primary datasets are unavailable.
-Furthermore even with primary data in hand, data reduction to effect sizes within primary and synthesis studies is a mechanism to illustrate differences and strength of effects.
-These approaches provide the capacity for higher-order analyses and reuse [@RN4835] suggesting that familiarity with effect sizes is both germane and practical.
-
-### The R ecosystem for meta-analyses  
-Like many fundamental challenges in science, the R developer community provides potential solution sets distributed across multiple packages for synthesis.
-Broadly speaking, alternative packages in R sometimes examine an issue from different perspectives and provide unique functions.
-In other instances, packages can be very similar or analogs in terms of functionality and use conceptually aligned functions that differ only in nomenclature or arguments [@RN6507].
-Scientific synthesists that choose to do a meta-analyses in R have options.
-A total of 63 packages associated with various aspects of conducting a meta-analysis have been identified in a comprehensive review and typology of options [@RN6178].
-Both meta [@RN6176] and metafor [@RN6175] are amongst 11 generic packages identified [@RN6178].
-These two packages are analogs but with different inherent workflows.
-There is also rmeta for simple fixed and random effects meta-analyses [@RN6195], mada for diagnostics [@RN6196], [@RN6196], netmeta for frequentist network meta statistics [@RN6197], and mvmeta for multivariate derived data aggregations [@RN6198] to name a few options.
-The latter three packages listed have distinct and specific niches for analysis whilst meta and metafor overlap considerably.
-Consequently, a brief contrast in facilitating choice between these two packages for the general analyst is provided here.
-
-### Contrast of meta versus metafor  
-Meta is a well-maintained, recently updated CRAN R package (Version 4.11-0 updated on Feb 20, 2020) with 31 unique functions, 7 sample datasets (Appendix A), and a reference manual.
-There is also an exceptional textbook devoted to meta-analysis in R that focuses primarily on this package [@RN6199].
-It is highly capable of resolving most general meta-analytical challenges that an analyst will face including the capacity to include Empirical Bayes estimators as arguments in some functions, predictive meta-statistics, interaction terms, meta-regression, and modifiers.
-Note that for some of these methods, the rma.uni function is sourced internally from the package metafor.
-This is intriguing but mostly opaque and inconsequential to the user if she prefers the structure of the arguments within functions, the semantics, or the workflow of the meta package.
-The primary strengths include its direct and straightforward implementation with minimal (source) lines of code to do an analysis.
-Provided one has secured the derived data from the studies and organized into a dataframe with vectors as each key argument within the main meta-model fitting functions, statistics are simple.
-The type of response variable such as mean, continuous, or rate is matched to a specific function call such as metamean, metacont, or metarate.
-This is semantically intuitive and encourages good thinking before statistics because it engenders consideration of the data.
-The effect size calculation is included in this main function and defaults return the most prevalent effect size measure typically associated with those data, but it can also be specified as an argument.
-The primary workflow is thus a single step if the user elects to rely on the internal calculations provided in this package.
-Exploration of the model is well articulated with funnel, radial, and forest plots.
-Z-scores, significance tests, and heterogeneity statistics are printed in the model summary.
-Publication bias is also provided as a more in-depth function entitled metabias within this package.
-There are two standout functions in this package.
-The first is a function entitled metagen, and it is a backup, multipurpose tool so to speak that fits a generic inverse variance meta-analysis.
-This a handy tool for user-calculated effect size measures or for exploration of statistical trends with reduced data assumptions.
-In some fields, there are specific effect size estimates that this function provides a robust, easy-to-fit capacity for statistics.
-The second standout function is bubble.metareg for a quick, visual exploration of the outcome of a meta-regression.
-It is useful in contemporary data science to use visualization as a means to understand data [@RN4510], but statistical packages do not always provide the means to easily iterate between statistics or model fitting and visualization.
-In summary, excepting unique data or statistical issues, this package is directly implemented and effective.
-
-Metafor is a more comprehensive package in many respects.
-This package includes 74 functions, 35 datasets (Appendix A), a vignette [@RN4896], flowchart as secondary vignette (https://cran.r-project.org/web/packages/metafor/vignettes/diagram.pdf), and website (http://www.metafor-project.org/doku.php).
-The package was last updated on March 19, 2020 (Version 2.4-0).
-The text ‘Meta-analysis with R’ also describes implementation of this package [@RN6199] but to a lesser extent than meta.
-The depth of the package metafor provides greater capacities relative to the meta package but does come at the expense of a steeper initial learning curve.
-Completing a meta-analysis using this package requires an additional step, i.e.
-effect sizes must be calculated a priori, not within the model fitting process.
-This is facilitated with the standalone function escalc, and it can return a wide range of effect sizes measures.
-Thus, the two-step process begins with firstly compiling and aggregating the derived dataframe to an effect size table then secondly fitting a model.
-The data structure is also a bit more rigid for the model fitting, and the nomenclature for this subset of functions is written to parallel more traditional general linear model fitting from conventional statistics.
-This is both a strength and limitation because one must plan the model to fit in advance and learn the function and arguments, but it is also an advantage as well because model specification uses the familiar notation of tilde.
-Model fitting is based on the type of model in the call such a random or fixed effects and not on the type of the response data as in meta package.
-Here, it is more akin to conventional general linear model fitting for those familiar with these functions in R.
-If the model is more complex with moderators, then this can be directly included in the model fit here via a mods argument whereas in the meta package the model is updated with moderators in a subsequent step.
-This suggests that if moderators or covariates in the main model are likely relevant to the analyses, then metafor is a strong starting point.
-The model summary also prints Z-scores, significance tests, and two sets of heterogeneity estimates.
-Forest and radial plots are also provided as additional functions.
-Publication bias statistical estimator functions include trimfill and ranktest.
-Standout elements of this package include GOSH plots that provide a graphical display of study heterogeneity [@RN6582] and the enhanced model fitting capacities such as the function fitstats that provides log-likelihood estimates and AIC or BIC scores on meta-analysis objects.
-This package requires a focus on model fitting, and while there is additional effort in specifying the data at the onset of the workflow, the rewards in subsequent tools to handle models are significant.
-
-### Conclusions  
-Statistics are sometimes about preferences and thinking styles [@RN6087], and scientific synthesis is both an art and a science [@RN3629].
-Trade-offs are also common in adopting one ecosystem, analysis tool, or specific package for data wrangling and analyses.
-If more rapid, less specified, general meta-analyses are the goal – the package meta is a direct means to an end.
-Moderators are added post hoc in additional, update model steps, but the first model fit is a single, intuitive process.
-Meta-regression is viable and interaction terms can be included.
-The generic meta-analysis function is a superb tool.
-Metafor requires the effect size compilation a priori and is thus a bit more coding to prepare for the meta-model.
-However, deeper and more complex model fits are inherent in the semantics of these functions.
-If the synthesist does have not effect size measure in hand or wishes to calculate effect sizes measures but not for meta-models, the escalc function is invaluable in this package.
-In summary, both packages provide the capacity for basic and advanced meta-analyses but more advanced modelling is likely worth the commitment to metafor.
-
-\newpage
-### Table 1.  
-A contrast of meta and metafor using
-
-\newpage
-### Appendix A.  
 A list of objects included in each package when loaded into the R environment.  
@@
 ```
-
-\newpage
-### Literature cited

### 2020-03-30 ms/2020/ms.Rmd
getting there.

--- 2020-03-30 ms/2020/contrastRmd.Rmd
@@
 \newpage
-### Appendix A.  
-A list of objects included in each package when loaded into the R environment.  
-```{r appendix, message=FALSE, warning=FALSE, echo=FALSE}
-library(tidyverse)
-library(kableExtra)
-
-appendix <- read_csv("appendix.csv")
-knitr::kable(appendix, "latex", booktabs = TRUE, longtable = TRUE) %>%
-  column_spec(1, width = "2cm") %>%
-  column_spec(2, width = "4cm") %>%
-  column_spec(3, width = "5cm") %>%
-  column_spec(5, width = "2cm") %>%
-  kable_styling(latex_options = c("hold_position", "repeat_header"))
-```
-
-\newpage
 ### Literature cited

### 2020-03-30 contrast.Rmd
done contrasts

@@
 ---
-title: ""
+title: "Appendix B"
 output:
@@
 
-### A contrast of meta and metafor
+### A contrast of analyses done using the R packages meta and metafor.  
 Contrasted meta and metafor using the classic [Cochrane bronchoconstriction dataset]((https://www.cochrane.org/CD001183/AIRWAYS_nedocromil-sodium-for-preventing-exercise-induced-bronchoconstriction)) used in many texts to demonstrate meta-analsysis.
@@
 library(tidyverse) #tidy functions to read and format data
-library(meta) #meta package current version
+library(meta) #meta package current version effective March 2020
 
@@
 radial(m1)
-
-#cumulative evidence
-forest(metacum(m1, pooled = "random"))
-
 ```
 
+\newpage
 ### Metafor
@@
 library(tidyverse) #tidy functions to read and format data
-library(metafor) #metafor package current version
+library(metafor) #metafor package current version effective March 2020
 
@@
 
+#calculate effect sizes
+data <- escalc(measure = "SMD", m1i=Me,m2i=Mc,sd1i=Se,sd2i=Sc,n1i=Ne,n2i=Nc, data=data,var.names=c("SMD","SMD_var"),digits=4) #"SMD" for the standardized mean difference.
+
+#fit model
+#m2 <- rma(yi=SMD, vi=SMD_var, mods = ~study.ID, data = data)
+m2 <- rma.uni(yi=SMD, vi=SMD_var, data = data, method = "REML") #restricted maximum-likelihood estimate
+summary(m2)
+plot(m2)
+
 ```

### 2020-03-31 contrast.Rmd
app b done

@@
 
-#### Meta
+### Meta
 ```{r meta, warning=FALSE, message=FALSE}
@@
 summary(m1)
+
+#viz
 forest(m1, layout = "JAMA")
@@
 summary(m2)
+
+#Viz
 plot(m2)
+```
+
+\newpage
+### Compiled contrasts
+```{r contrasts, warning=FALSE, message=FALSE}
+#compile random model outputs for contrast with metafor
+estimate <- as_data_frame(m1$TE.random)
+se <- as_data_frame(m1$seTE.random)
+zval <- as_data_frame(m1$zval.random)
+pval <- as_data_frame(m1$pval.random)
+ci.lb <- as_data_frame(m1$lower.random)
+ci.ub <- as_data_frame(m1$upper.random)
+m1_tidy <- bind_cols(estimate, se, zval, pval, ci.lb, ci.ub) %>%
+  rename(estimate = value, se = value1, zval = value2, pval = value3, ci.lb = value4, ci.ub = value5)
+
+#compile output into table for contrast with meta
+m2_tidy <- coef(summary(m2))
+#m2_tidy
+
+#compile key outputs
+contrast_tidy <- bind_rows(m1_tidy, m2_tidy) %>%
+  mutate(source = c("meta", "metafor")) %>%
+  select(source, everything())
+
+knitr::kable(contrast_tidy)
 

### 2020-03-31 ms/2020/appendix_a.Rmd
done contrasts
app a updated - thanks Alex

@@
 
-A list of objects included in each package when loaded into the R environment.  
+### A list of objects included in each package when loaded into the R environment.
+
 ```{r appendix, message=FALSE, warning=FALSE, echo=FALSE}
@@
 
-appendix <- read_csv("appendix.csv")
+appendix <- read_csv("appendix_2020.csv")
 knitr::kable(appendix, "latex", booktabs = TRUE, longtable = TRUE) %>%

### 2020-03-31 ms/2020/ms.Rmd
app a updated - thanks Alex

@@
 
+Models are powerful and typically necessary tools to better understand data.
+Conservatively, a model is a mechanism to explore uncertainty and assign weight to an observed pattern in the data [@RN6117}.
+Any statistical test albeit simple is thus a model to help describe a trend and difference.
+Less generously, it has been proposed that all models are wrong but that some are useful [@RN6467].
+In this vein, it is not unreasonable to examine the outputs of similar packages to
+
 ### Conclusions  

### 2020-03-31 ms/2020/ms.Rmd
done!

--- 2020-03-30 ms/2020/ms.Rmd
@@
 Nonetheless, metafor has a steeper learning curve but greater rewards.
+Reference to the learning curve and capacities of the statistical software Stata is also provided as a benchmark outside the R ecosystem.
 
@@
 
-### Contrast of meta versus metafor  
-Meta is a well-maintained, recently updated CRAN R package (Version 4.11-0 updated on Feb 20, 2020) with 31 unique functions, 7 sample datasets (Appendix A), and a reference manual.
+### Contrasts  
+Meta is a well-maintained, recently updated CRAN R package (Version 4.11-0 updated on Feb 20, 2020) with 33 unique functions, 7 sample datasets (Appendix A), and a reference manual.
 There is also an exceptional textbook devoted to meta-analysis in R that focuses primarily on this package [@RN6199].
@@
 Metafor is a more comprehensive package in many respects.
-This package includes 74 functions, 35 datasets (Appendix A), a vignette [@RN4896], flowchart as secondary vignette (https://cran.r-project.org/web/packages/metafor/vignettes/diagram.pdf), and website (http://www.metafor-project.org/doku.php).
+This package includes 76 functions, 36 datasets (Appendix A), a vignette [@RN4896], flowchart as secondary vignette (https://cran.r-project.org/web/packages/metafor/vignettes/diagram.pdf), and website (http://www.metafor-project.org/doku.php).
 The package was last updated on March 19, 2020 (Version 2.4-0).
@@
 
+Models are powerful and typically necessary tools to better understand data.
+Conservatively, a model is a mechanism to explore uncertainty and assign weight to an observed pattern in the data [@RN6117].
+Any statistical test, albeit simple, is thus a model to help describe a trend or difference.
+Less generously, it has been proposed that all models are wrong but that some are useful [@RN6467].
+In this vein, it is not unreasonable to examine the outputs of similar packages to ensure that reported findings are relatively consistent [@RN6507].
+This is also an important consideration that can better advance replication science, i.e., can we repeat analyses within our fields for the exact same data using different tools and reach similar conclusions [@RN6481].
+A Cochrane dataset commonly used in teaching and many texts on meta-analyses [@RN6585] was tested in both meta and metafor to ensure that outputs were similar (Appendix B).
+The reported statistics from simple univariate, random models were nearly identical from these two R packages.
+The purpose of this exercise is to illustrate that it is worthwhile to consider adding contrasts to a meta-analytical workflow.
+This worked example also provides a concrete instance of the differences in coding and semantics associated with the implementation of each package described above.
+However, the intent here was not to comprehensively examine sensitivity between these two options nor rigourously examine model similarities.
+We also do not mean to imply that these packages will always return similar outputs for more advanced models or for different data.
+If external validity and higher levels of certainty in the strength of findings and robustness of statistics are needed, Stata is an additional resource to triple-check your work (or if you do not work in R, an alternative ecosystem).
+It is a common application used in many disciplines such as the social sciences for meta-analyses.
+There are extensive resources to support meta-analyses in Stata including descriptions of workflows and worked examples [@RN6583].
+It was also recently updated, December 2019 (Version 16), includes 13 key functions for meta-analyses, a reference manual specific to meta-analyses [@RN6584], and adopts a workflow akin to metafor.
+Prepare you data and calculate effect sizes, run a meta-analysis on these effect sizes and inspect summary statistics, explore heterogenity, and check for small-study effects and publication bias.
+Tools are provided for each step.
+We used the same Cochrane data in Stata 16 and repeated the analysis (Appendix C, restricted maximum-likelihood estimate also specified).
+The reported estimates, p-value, and confidence limits were also nearly identical to both R packages in this instance.
+Collectively, this suggests that the synthesis scientist has many viable options and that checks between tools are within reach to explore sensitivity.
+
 ### Conclusions  
@@
 In summary, both packages provide the capacity for basic and advanced meta-analyses but more advanced modelling is likely worth the commitment to metafor.
+The primary purpose of this computational tools commentary was to contrast the two most common R packages, but we would be remiss without highlighting that R is not the only ecosystem that supports meta-analyses.
+There are other applications such as Stata that have the capacity to do meta-analyses and validate/generate models.
+Depending on your analytical workflows and the need to validate findings, the learning curve in Stata is relatively shallow.
+Replication science is course course best realized through documented, published, and open code, but there are choices with different relative strengths.
 

### 2020-04-01 downloads.Rmd
r package downloads completed

---
title: "downloads"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Package use
### Data
Grab the data from CRAN.

```{r data, warning=FALSE, message=FALSE}
library(tidyverse)
library(dlstats)
library(lubridate)

data <- cran_stats(c("meta", "metafor"))

#tidy up
data <- data %>%
  mutate(int = interval(ymd(start), ymd(end)), days = time_length(int, "day")) %>%
  select(days, everything(), -int)

#summary by net totals
totals <- data %>%
  group_by(package) %>% 
  summarise(total_downloads = sum(downloads), total_days = sum(days))

#summary by months
months <- data %>%
  mutate(month = month(start)) %>%
  group_by(month, package) %>% 
  summarise(total_downloads = sum(downloads), total_days = sum(days))

#write_csv(data, "data/downloads.csv")
#write_csv(totals, "data/total_downloads.csv")

```

### Viz
Check popularity.

```{r viz, warning=FALSE, message=FALSE}
ggplot(data, aes(package, downloads)) +
  geom_boxplot() +
  scale_color_brewer(palette = "Set1")

ggplot(months, aes(month, total_downloads, color = package)) +
  geom_point() +
  geom_smooth() +
  scale_color_brewer(palette = "Set1") +
  scale_x_continuous(limits = c(1, 12), breaks = seq(1,12))

```

### 2020-04-13 ms/2020/cover_letter.Rmd
2020 cover letter

---
title: ""
output:
  pdf_document: default
  html_document: default
---
April 13^th^, 2020

### Cover letter
### A contrast of meta and metafor packages for meta-analyses in R

Dear Editors,

Meta-analyses are critical scientific synthesis tools.  Their use is increasing exponentially in the natural sciences, and R is a frequent and logical choice to get the data processed and statistics done.  In many disciplines including ecology, evolution, and the environmental sciences in particular, two packages dominate the contemporary landscape. A contrast here is provided to guide the computational synthesist in choosing between these two options. Both excel but in different capacities. This contrast further highlights the concepts of workflows, semantics, and preference as a set of criteria for computational biologists to consider for any challenge given the incredible volume and range of packages available on CRAN for R users and in most other computational environments. To expand the breadth, we also include a contrast with Stata because it is an advanced statistical software choice used by some data scientists. It functions here as a benchmark or external check on the capacity for implementation and analysis by these two common R packages. This contribution is a Perspective or Editorial describing computational tool use for biologists. We are confident that it is of wide interest not just to R users or those that use Stata because it demonstrates the capacity for formal contrasts in computational choice to inform needs-driven data solutions in biology.

Best,  
Christopher Lortie and Alex Filazzola

### 2020-04-28 ms/2020/MEE_cover_letter.Rmd
MEE setup

---
title: ""
output:
  pdf_document: default
  html_document: default
---
April 28^th^, 2020

### Cover letter
### A contrast of meta and metafor packages for meta-analyses in R

Dear Editors,

These are challenging times. In addition to health and well-being issues, eeb researchers face limited capacities to do their field and lab research. Many have turned to scientific synthesis including meta-analysis. We provide a short commentary here to facilitate this process, and we are certain that it will be helpful to our community.  If you agree, we would love to get this out there rapidly in whatever format you decide is more appropriate.

Meta-analyses are critical scientific synthesis tools.  Their use is increasing exponentially in the ecology and evolution, and R is a frequent and logical choice to get the data processed and statistics done.  In many disciplines including ecology, evolution, and the environmental sciences in particular, two packages dominate the contemporary landscape. A contrast here is provided to guide the synthesist in choosing between these two options. Both excel but in different capacities. To expand the breadth, we also include a contrast with Stata because it is an advanced statistical software choice used by some data scientists. It functions here as a benchmark or external check on the capacity for implementation and analysis by these two common R packages. This contribution is a short commentary describing application and practical computational tools. We are confident that it is of wide interest not just to R users or those that use Stata because it demonstrates the capacity for formal contrasts between different methods (like we do in the field) to inform needs-driven solutions in ecology and evolution.

Best,  
Christopher Lortie and Alex Filazzola

### 2020-04-28 ms/2020/MEE_ms.Rmd
MEE setup

--- 2020-03-31 ms/2020/ms.Rmd
@@
 output: pdf_document
+header-includes:
+    - \usepackage{setspace}\doublespacing
+    - \usepackage[left]{lineno}
+    - \linenumbers
 bibliography: references.bib
@@
 3. Biological Sciences, University of Alberta,  
-Edmonton, AB, Canada,
+Edmonton, AB, Canada.
 
+Running headline: Meta-analysis application contrast.
+
+\newpage
 ### Abstract
-There is extensive choice in R to support meta-analyses.
-Two common packages used in the sciences include meta and metafor.
-A brief contrast of their strengths is described here for the synthesis scientist.
-Meta is a direct, intuitive choice for rapid implementation of general meta-analytical statistics.
-Metafor is a comprehensive package best suited for relatively more complex models.
+1. There is extensive choice in R to support meta-analyses.  Two common packages used in the sciences include meta and metafor.  
+2. A brief contrast of their strengths is described here for the synthesis scientist.  
+3. Meta is a direct, intuitive choice for rapid implementation of general meta-analytical statistics.  Metafor is a comprehensive package best suited for relatively more complex models.  
+4.
 Both packages provide estimates of heterogeneity, excellent visualization tools, and functions to explore publication bias.
@@
 
+### Author contributions
+CJL conceived the idea and wrote the manuscript. AF edited the manuscript. AF tested the case study in Stata. CJL tested the case study in R.
+
+### Data availability
+Data available from the Figshare Data Repository https://figshare.com/articles/Meta_thinking_data_preventing_exercise-induced_bronchoconstriction/7767020 (Lortie 2019).
+
+### Appendix A. A list of objects included in each package when loaded into the R environment.
+
+### Appendix B. A contrast of analyses done using the R packages meta and metafor.
+
+### Appendix C. An example of a meta-analysis conducted in Stata 16.
+
 \newpage

### 2020-06-30 ms/2020/ms.Rmd
appendices listed.
setup

@@
 title: "A contrast of meta and metafor packages for meta-analyses in R"
-output: pdf_document
+output:
+  word_document: default
+  pdf_document: default
 bibliography: references.bib
@@
 
+### Appendix A. A list of objects included in each package when loaded into the R environment.
+
+### Appendix B. A contrast of analyses done using the R packages meta and metafor.
+
+### Appendix C. An example of a meta-analysis conducted in Stata 16.
+
 \newpage
